---
title: "Alzheimer's disease"
author: "Ruben Cabrera"
date: "2024-07-19"
output: 
  github_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Nonlinear Mixed Effects Modeling

Based in https://github.com/larslau/progmod?tab=readme-ov-file


```{r intro}
# install.packages('devtools')
# devtools::install_github('larslau/progmod')

library(progmod)
library(ggplot2)
```

## Data

```{r data}

summary(adas_mmse_data)
str(adas_mmse_data)
head(adas_mmse_data)

# Expressing blstatus as factor (ordered):

adas_mmse_data$blstatus <- factor(adas_mmse_data$blstatus, levels = c("Cognitively normal", "MCI", "Dementia")) 
```

As you can see, we have the variables ADAS13 and MMSE, indicators of mental capacity.
The bigger ADAS13 number, the lower mental capacity
The lower MMSE scores, the lower mental capacity.


We have also predefined blstatus as:

-   Cognitively normal
-   MCI
-   Dementia

Let's explore the data with some plots





```{r ini_plots}
plot(adas_mmse_data$ADAS13,adas_mmse_data$MMSE)
plot(adas_mmse_data$Month_bl,adas_mmse_data$MMSE)

boxplot(adas_mmse_data$ADAS13 ~ adas_mmse_data$blstatus)
boxplot(adas_mmse_data$MMSE ~ adas_mmse_data$blstatus)

sampledata <- adas_mmse_data[adas_mmse_data$subject_id == 627 | adas_mmse_data$subject_id == 358 | adas_mmse_data$subject_id == 168,]

sampledata


## ADAS13 for sampledata

ggplot(sampledata, aes(x = Month_bl, y = ADAS13)) +
geom_line(aes(group = subject_id, color = blstatus)) +
ylim(c(85, 0)) + xlab('Months since baseline')

## MMSE for sampledata

ggplot(sampledata, aes(x = Month_bl, y = MMSE)) +
geom_line(aes(group = subject_id, color = blstatus)) +
ylim(c(0, 30)) + xlab('Months since baseline')

## ADAS13 for all data

ggplot(adas_mmse_data, aes(x = Month_bl, y = ADAS13)) +
geom_line(aes(group = subject_id, color = blstatus)) +
ylim(c(85, 0)) + xlab('Months since baseline')

## MMSE for all data

ggplot(adas_mmse_data, aes(x = Month_bl, y = MMSE)) +
geom_line(aes(group = subject_id, color = blstatus)) +
ylim(c(0, 30)) + xlab('Months since baseline')


```

## Model to simulated ADAS-cog scores

Fit exponential disease progression model to simulated ADAS-cog scores

```{r adas13model}
fixed_start_coef <- c(0.5, 70, 150, 3.5, 10)

ADAS_progmod <- progmod(ADAS13 ~ exp_model(Month_bl, l, s, g, v),
                     data = subset(adas_mmse_data, !is.na(ADAS13)),
                     fixed = list(l ~ 1,
                                  s ~ MCI + DEM - 1,
                                  g ~ 1,
                                  v ~ 1),
                     random = s + v ~ 1 | subject_id,
                     start = fixed_start_coef,
                     covariance = NULL)

summary(ADAS_progmod)

# Predict from model and visualize results
adas_mmse_data$fixed_shift_adas <- with(adas_mmse_data,
                                       MCI * fixed.effects(ADAS_progmod)[2] +
                                        DEM * fixed.effects(ADAS_progmod)[3])

pred_rand <- random.effects(ADAS_progmod)

adas_mmse_data$random_shift_adas <- pred_rand[match(adas_mmse_data$subject_id, 
                                                    rownames(pred_rand)), 's.(Intercept)']


sampledata <- adas_mmse_data[adas_mmse_data$subject_id == 627 | adas_mmse_data$subject_id == 358 | adas_mmse_data$subject_id == 168,]

sampledata

ggplot(sampledata, aes(x = Month_bl + fixed_shift_adas + random_shift_adas, y = ADAS13)) +
    geom_line(aes(group = subject_id, color = blstatus)) +
    ylim(c(85, 0)) +
    xlab('Months since baseline')

ggplot(adas_mmse_data, aes(x = Month_bl + fixed_shift_adas + random_shift_adas, y = ADAS13)) +
    geom_line(aes(group = subject_id, color = blstatus)) +
    ylim(c(85, 0)) +
    xlab('Months since baseline')

```


## Model to simulated MMSE scores

Fit generalized logistic model with range [30, 0] and a random time shift

```{r mmsemodel}
fixed_start_coef <- c(B = 0.025,
                      v = 1.4,
                      `s.(Intercept)` = -100,
                      s.MCI = 26,
                      s.DEM = 75)

MMSE_progmod_glf <- progmod(MMSE ~ GLF(Month_bl, A = 30, K = 0, B, v, s, c = 0),
                            data = subset(adas_mmse_data, !is.na(MMSE)),
                            fixed = list(B ~ 1,
                                         v ~ 1,
                                         s ~ MCI + DEM + 1),
                            random = s ~ 1 | subject_id,
                            start = fixed_start_coef,
                            covariance = NULL)

summary(MMSE_progmod_glf)

# Predict from model and visualize results
adas_mmse_data$fixed_shift_mmse <- with(adas_mmse_data,
                                        fixed.effects(MMSE_progmod_glf)[3] +
                                        MCI * fixed.effects(MMSE_progmod_glf)[4] +
                                        DEM * fixed.effects(MMSE_progmod_glf)[5])

pred_rand <- random.effects(MMSE_progmod_glf)

adas_mmse_data$random_shift_mmse <- pred_rand[match(adas_mmse_data$subject_id, 
                                                    rownames(pred_rand)), 's.(Intercept)']


sampledata <- adas_mmse_data[adas_mmse_data$subject_id == 627 | adas_mmse_data$subject_id == 358 | adas_mmse_data$subject_id == 168,]

sampledata

ggplot(sampledata, aes(x = Month_bl + fixed_shift_mmse + random_shift_mmse, y = MMSE)) +
    geom_line(aes(group = subject_id, color = blstatus)) +
    xlab('Months since baseline')

ggplot(adas_mmse_data, aes(x = Month_bl + fixed_shift_mmse + random_shift_mmse, y = MMSE)) +
    geom_line(aes(group = subject_id, color = blstatus)) +
    xlab('Months since baseline')

```

## Format

```{r format}

# Stack data to long format
tmp1 <- adas_mmse_data[, c('subject_id', 'Month_bl', 'CN', 'MCI', 'DEM', 'ADAS13')]

names(tmp1)[6] <- 'value'

tmp1$scale <- 'ADAS13'

tmp2 <- adas_mmse_data[, c('subject_id', 'Month_bl', 'CN', 'MCI', 'DEM', 'MMSE')]

names(tmp2)[6] <- 'value'

tmp2$scale <- 'MMSE'

# Long data
adas_mmse_data_long <- na.omit(rbind(tmp1, tmp2))

adas_mmse_data_long$scale <- factor(adas_mmse_data_long$scale)

# Remove temporary files
rm(tmp1, tmp2)

# Fit multivariate exponential model
fixed_start_coef <- c(l.scaleADAS13 = 0.5,
                      l.scaleMMSE = -0.1,
                      s.MCI = 70,
                      s.DEM = 150,
                      g.scaleADAS13 = 3.5,
                      g.scaleMMSE = 3,
                      v.scaleADAS13 = 10,
                      v.scaleMMSE = 30)



multi_progmod_glf <- progmod(value ~ exp_model(Month_bl, l, s, g, v),
                             data = adas_mmse_data_long,
                             fixed = list(l ~ scale + 0,
                                          s ~ MCI + DEM + 0,
                                          g ~ scale + 0,
                                          v ~ scale + 0),
                             random = list(s ~ 1,
                                           v ~ scale),
                             groups = ~ subject_id,
                             start = fixed_start_coef,
                             weights = varIdent(form = ~ 1 | scale))

# Predict from model and compare to univariate models
adas_mmse_data$fixed_shift_multi <- with(adas_mmse_data,
                                         MCI * fixed.effects(multi_progmod_glf)[3] +
                                           DEM * fixed.effects(multi_progmod_glf)[4])

pred_rand <- random.effects(multi_progmod_glf)

adas_mmse_data$random_shift_multi <- pred_rand[match(adas_mmse_data$subject_id, 
                                                     rownames(pred_rand)), 's.(Intercept)']

# Correlations between predicted disease months
with(adas_mmse_data, cor(cbind(fixed_shift_adas + random_shift_adas,
                               fixed_shift_mmse + random_shift_mmse,
                               fixed_shift_multi + random_shift_multi), method = 'spearman'))
 
 




```